Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-28 src/article.tex
Revision to results and discussion section

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple format to write
  and edit scaffold file syntax. This syntax further allows  unknown regions to
  be defined, and inserts to be added to the scaffold to fill gaps. Defining
  the genome construction in a file makes the scaffolding process reproducible
  and easier to edit and compare than fasta sequence.

@@

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence inserted from the corresponding fasta file. The
Scaffolder software processes this list in order, converting each to the
corresponding nucleotide sequences, which are then concatenated to produce
a draft super sequence. Each entry in the scaffold file may have attributes
that define whether a sub-sequence or the reverse complement of the sequence
should be used. The types of attributes available, and an example scaffold file
are outlined below. \pb

@@

The Scaffolder software outlined here allows computational joining of
nucleotide sequences together into a single contiguous super sequence.
Plain-text scaffold files written in the YAML syntax specify how these
sequences should be joined. This scaffold file is read by the scaffolder
software and used to generate the scaffold sequence. In addition to specifying
which contigs are required, the scaffold file allows the contigs to be edited
to produce smaller sub-sequences or reverse complemented if necessary. \pb

The process of genome finishing may involve producing additional
oligonucleotide sequence to fill unknown regions in the scaffold. The scaffold
file format allows these additional insert sequences to be added to bridge
these gaps. These inserts can also be treated in a similar way to larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region. \pb

Distances between contigs may be estimated from paired read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join regions of sequence together by an estimated size. The scaffold
file allows for the specification of these unknown regions to join contigs
together. The use of these regions in the scaffold highlights areas still to be
resolved. \pb

The scaffold file is used to specify how sequence regions are joined together
but the sequences themselves are not stored in the scaffold file. The sequences
are instead maintained as a separate fasta file. A fasta file of contigs and
scaffolded contigs is a common output for assemblers and as such should be
readily available. The nucleotide sequences in the fasta file are referenced in
the scaffold using the first space-delimited word from each fasta header.
Maintaining the nucleotide sequences in a separate file preserves the original
sequence and furthermore decouples the sequence from the markup describing how
it should be used to generate the genome sequence. \pb

@@

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a scaffold and the joining of sequences in the genome build.
The basic layout of the scaffold file is a list - a series of entries where
each entry corresponds to a region of sequence in the generated scaffold
super-sequence. The right hand side indicates how the entries are joined
together. This figure is illustrates the putative scaffold and is not to scale.
\pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. The first
entry, highlighted in green, in the scaffold begins on the second line. The
first line of this entry begins with a dash character `-', dashes are used to
denote entries in a YAML list. The first line of each entry in the scaffold
therefore begins with a dash line. The next line is indented by two spaces
- whitespace is required to group similar attributes together. \pb

The ``sequence'' tag indicates this entry corresponds to a sequence in the
fasta file. The following line indicates the name of the fasta sequence using
the ``source'' tag. As described above the first space-delimited word of the
fasta header is used to identify which sequence is selected from the fasta
file. Together these three lines describe the first entry in the scaffold as
a simple sequence using the fasta sequence named `sequence1'. On the right hand
side of Figure 1 this produces the first region in the scaffold, also shown in
green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the fasta entry
`sequence2' to the scaffold. This entry demonstrates how the nucleotide
sequence may be manipulated prior to addition to the scaffold. The `start' and
`stop' tags trim the sequence to these coordinates inclusively. The ``reverse''
tag instructs scaffolder to also reverse complement the sequence. On the right
hand side of the sequence of Figure 1 this completes the scaffold sequence. \pb

This final entry also includes the ``inserts'' tag to add additional regions of
sequence. These inserts are also added as a YAML list with each insert
beginning with a dash line. The first insert, shown in purple, illustrates
using similar attributes to that of a sequence entry. The reverse, start and
stop tags, are used to trim and reverse complement the insert. Similarly the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert fasta sequence. This is
highlighted in the putative scaffold in Figure 1 by the black lines bisecting
the blue sequence. \pb

The next insert, shown in brown, is specified using only the open position tag.
This illustrates that only one of either `open' or `close' tags is required.
When only one of these tag is used the corresponding opposing coordinate
position is calculated from the length of the insert fasta sequence. This
allows inserts to bridge into, and partially fill gap regions without requiring
an end coordinate position. \pb

@@

The the scaffold file format allows multiple sequence-editing operations to be
applied to a sequence or insert. If applied in different combinations, these
operations will produce different output sequences. An example would be reverse
complementing the sequence then trimming the start of the sequence by five
nucleotides. Performing these actions in the reverse order would result in five
nucleotides being removed from the opposite end. This section therefore
outlines the Scaffolder protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts, these inserts are reverse sorted by the insert close coordinate
position. The insert with the largest close position is first in the list and
the insert with the smallest close position is last. This preserves the
original intended coordinates as each insert is added to the sequence. \pb

\paragraph{Update sequence with inserts:} Each insert sequence is processed
according to any specified attributes. The insert is then added to the
enclosing sequence in the region specified by the open and close coordinates.
\pb

If the insert added to the sequence is a different size to that of the
replaced region then the sequence start and stop co-ordinates are updated. For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position the remaining
inserts are unaffected by changes in the encoding sequence size. \pb

\paragraph{Trimming sequence:} If the sequence has start or stop tags
specified these are used as coordinates to trim the sequence to the specified
length inclusively. \pb

\paragraph{Reverse complement sequence:} If the reverse tag is specified the
sequence is reverse complemented. This is final step in processing the
sequence and the sequence is then returned for inclusion in the scaffold. \pb

@@

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file
and the fasta file to be passed as arguments. This will then return the
complete scaffold sequence in fasta format. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable array and the penultimate line illustrates
the trivial code required to generate the super-sequence. Further details of
the classes and methods related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class in the documentation on the
Scaffolder website. \pb

@@

The aims of Scaffolder are twofold. The first is providing software which is
easy to install and simplifies the task of genome finishing. The second aim is
to facilitate reproducibility when joining large nucleotide sequences together
to create a genome sequence. \pb

\subsubsection*{Simple to use} %{{{3

Scaffolder is designed to be as simple to use as possible. Assuming the Ruby
and RubyGems software are present, Scaffolder can be installed in a single
command line step. This should negate any possible barriers to entry which
require manual compilation of source code. \pb

Scaffolder uses a minimal and compact syntax to describe the genome scaffold.
This syntax is simple to construct and edit whilst being succinct and readable.
The goal of this syntax is to make genome scaffolds easy to construct in
a common text editor. Furthermore Scaffolder provides extensive documentation
both for the CLI and for the Ruby API. Therefore should any part of the
scaffolding process be unclear the documentation should make an solution easy
to find. \pb

\subsubsection*{Reproducible genome construction} %{{{3

Constructing a genome sequence using a text file makes generating a scaffold
super sequence both reproducible and deterministic. In comparison joining large
nucleotide sequences by hand cannot be reliably reproduced. The layout of the
scaffold file furthermore provides a human readable record of how the scaffold
was constructed and decouples the layout of contigs from their source sequence.
\pb

>>>>

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple format to write
  and edit scaffold file syntax. This syntax further allows  unknown regions to
  be defined, and inserts to be added to the scaffold to fill gaps. Defining
  the genome construction in a file makes the scaffolding process reproducible
  and easier to edit and compare than FASTA sequence.

@@

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence inserted from the corresponding FASTA file. The
Scaffolder software processes this list in order, converting each to the
corresponding nucleotide sequences, which are then concatenated to produce
a draft super sequence. Each entry in the scaffold file may have attributes
that define whether a sub-sequence or the reverse complement of the sequence
should be used. The types of attributes available, and an example scaffold file
are outlined below. \pb

@@

The Scaffolder software allows computational joining of nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in the YAML syntax specify how these sequences should be
joined and the scaffold file as read by the scaffolder software is used to
generate the scaffold sequence. In addition to specifying which contigs are
required, the scaffold file allows the contigs to be edited to produce smaller
sub-sequences or reverse complemented if necessary. The process of genome
finishing may involve producing additional oligonucleotide sequences to fill
unknown regions in the scaffold. The scaffold file format allows these
additional insert sequences to bridge these gaps. These inserts can also be
treated in a similar way to larger contig sequences: trimmed and/or reverse
complemented to match the corresponding gap region. \pb

The distances between contigs may be estimated from paired read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join separate sequences together by an estimated size and the
scaffold file allows for the specification of these unknown regions. The use of
these regions in the scaffold which areas in the scaffold and thier approximate
size are still to be resolved. \pb

The scaffold file specifies how sequence regions are joined together; however
the sequences themselves are not stored in the scaffold file and are instead
maintained as a separate FASTA file. A FASTA file of contigs and scaffolded
contigs is a common output for assemblers. The nucleotide sequences stored in
the corresponding FASTA file are referenced in the scaffold using the first
word in the FASTA header. Maintaining the nucleotide sequences in a separate
file preserves the original sequence and decouples the original sequence data
from the scaffold file describing how it should be used to in the genome
sequence. \pb

@@

The scaffold file is written using the YAML syntax. An example scaffold file is
shown in Figure 1. This file illustrates the text attributes used to describe
a scaffold and the joining of sequences in the genome build. The basic layout
of the scaffold file is a list of entries, where each entry corresponds to
a region of sequence in the generated scaffold super-sequence. \pb

@@

The first line of the scaffold file begins with three dashes to indicate the
start of a YAML-formatted document. The first entry (highlighted in green)
begins with a dash character `-' to denote an entry in the YAML list. The first
line of each entry in the scaffold therefore begins with a dash line. The next
line is indented by two spaces as whitespace is required to group similar
attributes together. The ``sequence'' tag indicates that this entry corresponds
to a sequence in the FASTA file and the following line indicates the name of
this FASTA sequence using the ``source'' tag. As described the first
space-delimited word of the FASTA header is used to identify which sequence is
selected from the FASTA file. Together these three lines describe the first
entry in the scaffold as a simple sequence using the FASTA sequence named
`sequence1'. On the right hand side of Figure 1 this produces the first region
in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the FASTA entry
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively, while the ``reverse'' tag
instructs scaffolder to also reverse complement the sequence. In the scaffolded
genome shown in Figure 1 this completes the scaffold sequence. \pb

The final entry in the scaffold also includes the ``inserts'' tag to add
additional regions of sequence. These inserts are also added as a YAML list,
with each insert beginning with a dash line. The first insert, shown in purple,
similar attributes to that of a sequence entry; the reverse, start and stop
tags are used to trim and reverse complement the insert. Similarly the `source'
tag identifies the corresponding FASTA sequence as `insert'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert FASTA sequence. This is
visualised in the putative gene Figure 1 by the black lines bisecting the blue
sequence. \pb

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either `open' or `close' tags is required when
adding an insert sequence. If only only one of the `open' or `close' tags is
used the corresponding opposing position distance is calculated from the length
of the insert FASTA sequence. This allows inserts to bridge into, and partially
fill gap regions without requiring an end coordinate position. \pb

@@

The the scaffold file format allows multiple sequence-editing operations to be
applied to a sequence or insert. If applied in different combinations, then
these operations will produce different output sequences. An example would be
reverse complementing the sequence then trimming the start of the sequence by
five low base-call quality nucleotides. Performing these actions in the reverse
order would result in five nucleotides being removed from the opposite end.
This section therefore outlines the Scaffolder protocol for processing sequence
entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts, these inserts are sorted by decesding value of the insert `close' tag
value. The insert with the furthest `close' coordinate is first in the list and
the insert with the smallest close position is last. This preserves the
intended location for each insert if the size of the encoding sequence is
changed by the addition of inserts. \pb

\paragraph{Update sequence with inserts:} Each insert sequence is processed
according to any specified attributes, before being added to the enclosing
sequence in the region specified by the open and close coordinates. If the
insert added to the sequence is a different size to that of the replaced region
then the sequence start and stop coordinates are updated. For example if a 7bp
insert replaces a 5bp region in the encoding sequence, then the stop coordinate
of the enclosing sequence would be incremented by the 2bp difference. As
inserts are added by reverse close position the remaining inserts are
unaffected by changes in the encoding sequence size. \pb

\paragraph{Trimming sequence:} If the sequence has start or stop tags
specified, then these are used as coordinates to trim the sequence to the
specified length inclusively of the `start' and `stop' coordinate tags. \pb

\paragraph{Reverse complement sequence:} If the reverse tag is specified, then
the sequence is reverse complemented, which is the final is the final step in
processing the sequence. \pb

@@

The ``sequence'' command builds the FASTA sequence from the scaffold and FASTA
file. Running this command requires the locations of both the scaffold file and
the FASTA file to be passed as arguments. This will then return the complete
scaffold sequence in FASTA format. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the FASTA sequence
file. The scaffold is an enumerable array and the penultimate line illustrates
the trivial code required to generate the super-sequence. Further details of
the classes and methods related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class in the documentation on the
Scaffolder website. \pb

@@

The aims of Scaffolder are twofold. The first is providing software that is
easy to install and simplifies the task of genome finishing. The second aim is
to facilitate reproducibility when joining large nucleotide sequences together
to create a genome sequence. \pb

Scaffolder was designed to be as simple to use as possible and assuming the
Ruby and RubyGems software are present, Scaffolder can be installed in a single
command line step. This should negate any possible barriers to entry which
require manual compilation of source code. Scaffolder also uses a minimal and
compact syntax to describe the genome scaffold. This syntax is simple to
construct and edit whilst being succinct and readable. The goal of this syntax
is to make genome scaffolds easy to construct in a common text editor. \pb

Constructing a genome using a text file makes generating a scaffold super
sequence both reproducible and deterministic for the same scaffold and FASTA
sequence file. In comparison, joining large nucleotide sequences by hand cannot
be reliably reproduced. The scaffold file furthermore provides a human readable
description of how the scaffold is constructed. \pb
